---
title: "single cell"
author: "Ieva Gudelytė"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(Seurat)
library(Matrix)
library(patchwork)
library(dplyr)
library(ggplot2)
```

```{r}
options(Seurat.object.assay.version = "v3")
```

```{r}
counts <- readMM("counts.mtx.gz")
meta <- read.table("metadata.tsv.gz")
features <- read.table("features.tsv.gz")[,1]
dimnames(counts) <- list(features, rownames(meta))
seurat <- CreateSeuratObject(counts = counts, meta.data = meta)
```

```{r}
seurat_list <- SplitObject(seurat, split.by = "bio_sample")
```

```{r}
seurat_list <- lapply(seurat_list, function(x) {
  x[["percent.mt"]] <- PercentageFeatureSet(x, pattern = "^MT[-\\.]")
  x
})
```

```{r}
lapply(seurat_list, function(x) {
  VlnPlot(x, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
})
```

```{r}
lapply(seurat_list, function(x) {
  plot1 <- FeatureScatter(x, feature1 = "nCount_RNA", feature2 = "percent.mt")
  plot2 <- FeatureScatter(x, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
  plot1 + plot2
})
```

```{r}
seurat_list <- lapply(seurat_list, function(x) {
  x <- subset(x, subset = nFeature_RNA > 500 & nFeature_RNA < 3500 & percent.mt < 10)
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, nfeatures = 3000)
  x <- ScaleData(x)
  x <- ScaleData(x, vars.to.regress = c("nFeature_RNA", "percent.mt"))
  x
})
```

```{r}
seurat_list <- lapply(seurat_list, function(x) {
  x <- RunPCA(x, npcs = 50)
  x
})
```

```{r}
lapply(seurat_list, function(x) {
  ElbowPlot(x, ndims = ncol(Embeddings(x, "pca")))
})
```

```{r}
lapply(seurat_list, function(x) {
  PCHeatmap(x, dims = 1:10, cells = 500, balanced = TRUE, ncol = 4)
})
```

# Is data integration necessary?

```{r}
counts_test <- readMM("counts.mtx.gz")
meta_test <- read.table("metadata.tsv.gz")
features_test <- read.table("features.tsv.gz")[,1]
dimnames(counts_test) <- list(features_test, rownames(meta_test))
seurat_test <- CreateSeuratObject(counts = counts_test, meta.data = meta_test)
```

```{r}
seurat_list_test <- SplitObject(seurat_test, split.by = "bio_sample")
```

```{r}
seurat_test <- merge(x = seurat_list_test$Org25, y = c(seurat_list_test$Org28, seurat_list_test$Org56, seurat_list_test$Org54), add.cell.ids = c("Org25","Org28","Org56","Org54")) %>%
  NormalizeData() %>%
  FindVariableFeatures(nfeatures = 3000) %>%
 ScaleData() %>%
 RunPCA(npcs = 50) %>%
 RunUMAP(dims = 1:20)
```

```{r}
plot1 <- DimPlot(seurat_test, group.by="bio_sample")
plot2 <- FeaturePlot(seurat_test, c("FOXG1","EMX1","NSG1","SOX2"), ncol=2, pt.size =
0.1)
plot1 + plot2 + plot_layout(widths = c(1.5, 2))
```

Seems like it is.

# Data integration using Seurat

set dir
libraries
seurat_int_60 = readRDS("integrated_seurat_60.rds")
seurat_int_120 = readRDS("integrated_seurat_120.rds")

```{r}
counts_int <- readMM("counts.mtx.gz")
meta_int <- read.table("metadata.tsv.gz")
features_int <- read.table("features.tsv.gz")[,1]
dimnames(counts_int) <- list(features_int, rownames(meta_int))
seurat_int <- CreateSeuratObject(counts = counts_int, meta.data = meta_int)
```

```{r}
seurat_list_int <- SplitObject(seurat_int, split.by = "bio_sample")
```

```{r}
seurat_list_int <- lapply(seurat_list_int, function(x) {
  x[["percent.mt"]] <- PercentageFeatureSet(x, pattern = "^MT[-\\.]")
  x
})
```

```{r}
seurat_list_int <- lapply(seurat_list_int, function(x) {
  x <- subset(x, subset = nFeature_RNA > 500 & nFeature_RNA < 3500 & percent.mt < 10)
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, nfeatures = 3000)
  x
})
```

```{r}
seurat_list_int60 = seurat_list_int[c("Org25", "Org28")]
seurat_list_int120 = seurat_list_int[c("Org56", "Org54")]
anchors60 <- FindIntegrationAnchors(object.list = seurat_list_int60, dims = 1:30)
anchors120 <- FindIntegrationAnchors(object.list = seurat_list_int120, dims = 1:30)
```

```{r}
seurat_int_60 <- IntegrateData(anchors60, dims = 1:30)
seurat_int_120 <- IntegrateData(anchors120, dims = 1:30)
```

```{r}
seurat_int_60 <- ScaleData(seurat_int_60)
seurat_int_60 <- RunPCA(seurat_int_60, npcs = 50)
seurat_int_60 <- RunUMAP(seurat_int_60, dims = 1:30)
seurat_int_60 <- RunTSNE(seurat_int_60, dims = 1:30)
seurat_int_60 <- FindNeighbors(seurat_int_60, dims = 1:30) %>% FindClusters(resolution = 0.6)
```

```{r}
seurat_int_120 <- ScaleData(seurat_int_120)
seurat_int_120 <- RunPCA(seurat_int_120, npcs = 50)
seurat_int_120 <- RunUMAP(seurat_int_120, dims = 1:30)
seurat_int_120 <- RunTSNE(seurat_int_120, dims = 1:30)
seurat_int_120 <- FindNeighbors(seurat_int_120, dims = 1:30) %>% FindClusters(resolution = 0.6)
```

```{r}
DefaultAssay(seurat_int_60) <- "RNA"
plot1 <- UMAPPlot(seurat_int_60, group.by="bio_sample")
plot2 <- UMAPPlot(seurat_int_60, label = T)
plot3 <- FeaturePlot(seurat_int_60, c("SOX2","STMN2","HOPX","SYT1"), ncol = 2, pt.size = 0.1)
((plot1 / plot2) | plot3) + plot_layout(width = c(1,2))
```

```{r}
DefaultAssay(seurat_int_120) <- "RNA"
plot1 <- UMAPPlot(seurat_int_120, group.by="bio_sample")
plot2 <- UMAPPlot(seurat_int_120, label = T)
plot3 <- FeaturePlot(seurat_int_120, c("SOX2","STMN2","HOPX","SYT1"), ncol = 2, pt.size = 0.1)
((plot1 / plot2) | plot3) + plot_layout(width = c(1,2))
```

```{r}
saveRDS(seurat_int_60, file="integrated_seurat_60.rds")
saveRDS(seurat_int_120, file="integrated_seurat_120.rds")
```

```{r}
plot1 <- TSNEPlot(seurat_int_60, group.by="bio_sample")
plot2 <- UMAPPlot(seurat_int_60, group.by="bio_sample")
plot1 + plot2
```

```{r}
plot1 <- TSNEPlot(seurat_int_120, group.by="bio_sample")
plot2 <- UMAPPlot(seurat_int_120, group.by="bio_sample")
plot1 + plot2
```

```{r}
plot1 <- DimPlot(seurat_int_60, reduction = "umap", label = TRUE)
plot2 <- DimPlot(seurat_int_120, reduction = "umap", label = TRUE)
plot1 + plot2
```

```{r}
seurat_int_60[["RNA"]] <- as(seurat_int_60[["RNA"]], Class = "Assay")
class(seurat_int_60[["RNA"]])
seurat_int_120[["RNA"]] <- as(seurat_int_120[["RNA"]], Class = "Assay")
class(seurat_int_120[["RNA"]])
```

```{r, fig.width=37, fig.height=16}
DefaultAssay(seurat_int_60) <- "RNA"
DefaultAssay(seurat_int_120) <- "RNA"
ct_markers <- c("MKI67", "TOP2A", "PCNA", "CDK1", #cell cycle
"SOX2", "HES1", #dorsal forebrain progenitors 
"NEUROD2", "NEUROD6", #dorsal excitatory neurons
"GADD45G", "SSTR2", #neuronal migration 
"EMX1", "SOX2", "HES1", #neural progenitors, apical radial glia (aRG) 
"EMX1", "EOMES", "INSM1", #neural progenitors, intermediate progenitors (IP) 
"EMX2", "FOXP2", "EOMES", #the first neurons populating the developing cortical plate (preplate/subplate) 
"SOX2", "HOPX", "PEA15", "LGALS3BP", "MOXD1", #outer radial glia (oRG) 
"FEZF1", "NEUROD2", "NEUROD6", "TBR1", "PDE1A", #immature deep-layer projection neurons (DL PN) 
"SOX5", "LDB2", "CRYM", "TLE4", "BCL11B", "FEZF2", #deep-layer corticofugal projection neurons (CFuPN)
"UNC5A", "EPHA4", "BHLHE22", "PLXNA4", "SATB2", "CUX2", #callosal projection neurons (CPN) 
"GJA1", "S100B", "SPARC", "APOE", "AQP4", #glial cells with molecular features characteristic of astroglia
"OLIG1", "OLIG2", "PDGFRA", "EGFR", #oligodendrocyte precursors
"BIRC5", "DLX2", "DLX5", #cycling interneuron progenitors 
"GAD2", "DLX2", "DLX5", #immature interneurons 
"LAMP5", "LHX5", "GAD2", #subcortical interneurons
"PAX2", "CRABP1", "CRABP2", "TFAP2A", #subcortical interneurons, markers characteristic of caudal regions of the neural tube, i.e., hindbrain 
"FOXG1") #telencephalon / interneurons that appear after 4 months in culture
plot1 <- DoHeatmap(seurat_int_60, features = ct_markers) + guides(colour=FALSE)
plot2 <- DoHeatmap(seurat_int_120, features = ct_markers) + guides(colour=FALSE)
plot1 / plot2
```

```{r}
DefaultAssay(seurat_int_60) <- "RNA"
DefaultAssay(seurat_int_120) <- "RNA"
cl_markers_60 <- FindAllMarkers(seurat_int_60, only.pos = TRUE, min.pct = 0.25, logfc.threshold = log(1.2))
top10_cl_markers_60 <- cl_markers_60 %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)
cl_markers_120 <- FindAllMarkers(seurat_int_120, only.pos = TRUE, min.pct = 0.25, logfc.threshold = log(1.2))
top10_cl_markers_120 <- cl_markers_120 %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)
```

Cell cycle markers seem to have a significant effect on clustering. We want to get rid of this effect.
60 CCNB1 in cluster 10
60 MCM(10) in cluster 6
60 RRM2 in cluster 12
120 RRM2 in cluster 11

```{r}
DefaultAssay(seurat_int_60) <- "RNA"
DefaultAssay(seurat_int_120) <- "RNA"
seurat_int_60 <- CellCycleScoring(seurat_int_60,
 s.features = cc.genes$s.genes,
 g2m.features = cc.genes$g2m.genes,
 set.ident = TRUE)
seurat_int_120 <- CellCycleScoring(seurat_int_120,
 s.features = cc.genes$s.genes,
 g2m.features = cc.genes$g2m.genes,
 set.ident = TRUE)
```

```{r}
DefaultAssay(seurat_int_60) <- "RNA"
DefaultAssay(seurat_int_120) <- "RNA"
FeaturePlot(seurat_int_60, features = c("S.Score", "G2M.Score"))
FeaturePlot(seurat_int_120, features = c("S.Score", "G2M.Score"))
```

# Dealing with the cell cycle

```{r}
# DefaultAssay(seurat_int_60) <- "RNA"
# DefaultAssay(seurat_int_120) <- "RNA"
# VariableFeatures(seurat_int_60) <- setdiff(VariableFeatures(seurat_int_60), unlist(cc.genes))
# VariableFeatures(seurat_int_120) <- setdiff(VariableFeatures(seurat_int_120), unlist(cc.genes))
```

```{r}
# DefaultAssay(seurat_int_60) <- "integrated"
# DefaultAssay(seurat_int_120) <- "integrated"
# seurat_int_60 <- RunPCA(seurat_int_60, npcs = 50)
# seurat_int_60 <- RunUMAP(seurat_int_60, dims = 1:30)
# seurat_int_60 <- RunTSNE(seurat_int_60, dims = 1:30)
# seurat_int_60 <- FindNeighbors(seurat_int_60, dims = 1:30) %>% FindClusters(resolution = 1)
# seurat_int_120 <- RunPCA(seurat_int_120, npcs = 50)
# seurat_int_120 <- RunUMAP(seurat_int_120, dims = 1:30)
# seurat_int_120 <- RunTSNE(seurat_int_120, dims = 1:30)
# seurat_int_120 <- FindNeighbors(seurat_int_120, dims = 1:30) %>% FindClusters(resolution = 1)
```

```{r}
# plot1 <- UMAPPlot(seurat_int_60, group.by="bio_sample")
# plot2 <- UMAPPlot(seurat_int_120, group.by="bio_sample")
# plot1 + plot2
```

```{r}
# plot1 <- DimPlot(seurat_int_60, reduction = "umap", label = TRUE)
# plot2 <- DimPlot(seurat_int_120, reduction = "umap", label = TRUE)
# plot1 + plot2
```

```{r}
seurat_int_60[["RNA"]] <- as(seurat_int_60[["RNA"]], Class = "Assay")
class(seurat_int_60[["RNA"]])
seurat_int_120[["RNA"]] <- as(seurat_int_120[["RNA"]], Class = "Assay")
class(seurat_int_120[["RNA"]])
```

```{r}
# DefaultAssay(seurat_int_60) <- "RNA"
# DefaultAssay(seurat_int_120) <- "RNA"
# ct_markers <- c("MKI67", "TOP2A", "PCNA", "CDK1", #cell cycle
# "SOX2", "HES1", #dorsal forebrain progenitors 
# "NEUROD2", "NEUROD6", #dorsal excitatory neurons
# "GADD45G", "SSTR2", #neuronal migration 
# "EMX1", "SOX2", "HES1", #neural progenitors, apical radial glia (aRG) 
# "EMX1", "EOMES", "INSM1", #neural progenitors, intermediate progenitors (IP) 
# "EMX2", "FOXP2", "EOMES", #the first neurons populating the developing cortical plate (preplate/subplate) 
# "SOX2", "HOPX", "PEA15", "LGALS3BP", "MOXD1", #outer radial glia (oRG) 
# "FEZF1", "NEUROD2", "NEUROD6", "TBR1", "PDE1A", #immature deep-layer projection neurons (DL PN) 
# "SOX5", "LDB2", "CRYM", "TLE4", "BCL11B", "FEZF2", #deep-layer corticofugal projection neurons (CFuPN)
# "UNC5A", "EPHA4", "BHLHE22", "PLXNA4", "SATB2", "CUX2", #callosal projection neurons (CPN) 
# "GJA1", "S100B", "SPARC", "APOE", "AQP4", #glial cells with molecular features characteristic of astroglia
# "OLIG1", "OLIG2", "PDGFRA", "EGFR", #oligodendrocyte precursors
# "BIRC5", "DLX2", "DLX5", #cycling interneuron progenitors 
# "GAD2", "DLX2", "DLX5", #immature interneurons 
# "LAMP5", "LHX5", "GAD2", #subcortical interneurons
# "PAX2", "CRABP1", "CRABP2", "TFAP2A", #subcortical interneurons, markers characteristic of caudal regions of the neural tube, i.e., hindbrain 
# "FOXG1") #telencephalon / interneurons that appear after 4 months in culture
# plot1 <- DoHeatmap(seurat_int_60, features = ct_markers) + NoLegend()
# plot2 <- DoHeatmap(seurat_int_120, features = ct_markers) + NoLegend()
# plot1 + plot2
```

getting rid of cell cycle related genes from the identified highly variable gene list didn't seem to help much?

```{r}
DefaultAssay(seurat_int_60) <- "integrated"
DefaultAssay(seurat_int_120) <- "integrated"
seurat_int_60 <- ScaleData(seurat_int_60, vars.to.regress = c("S.Score", "G2M.Score"))
seurat_int_120 <- ScaleData(seurat_int_120, vars.to.regress = c("S.Score", "G2M.Score"))
```

```{r}
DefaultAssay(seurat_int_60) <- "integrated"
DefaultAssay(seurat_int_120) <- "integrated"
seurat_int_60 <- RunPCA(seurat_int_60, npcs = 50)
seurat_int_60 <- RunUMAP(seurat_int_60, dims = 1:30)
seurat_int_60 <- FindNeighbors(seurat_int_60, dims = 1:30) %>% FindClusters(resolution = 0.6)
seurat_int_120 <- RunPCA(seurat_int_120, npcs = 50)
seurat_int_120 <- RunUMAP(seurat_int_120, dims = 1:30)
seurat_int_120 <- FindNeighbors(seurat_int_120, dims = 1:30) %>% FindClusters(resolution = 0.6)
```

```{r, fig.width=37, fig.height=16}
Idents(seurat_int_60) <- "seurat_clusters"
Idents(seurat_int_120) <- "seurat_clusters"
DefaultAssay(seurat_int_60) <- "RNA"
DefaultAssay(seurat_int_120) <- "RNA"
ct_markers <- c("MKI67", "TOP2A", "PCNA", "CDK1", #cell cycle
"SOX2", "HES1", #dorsal forebrain progenitors 
"NEUROD2", "NEUROD6", #dorsal excitatory neurons
"GADD45G", "SSTR2", #neuronal migration 
"EMX1", "SOX2", "HES1", #neural progenitors, apical radial glia (aRG) 
"EMX1", "EOMES", "INSM1", #neural progenitors, intermediate progenitors (IP) 
"EMX2", "FOXP2", "EOMES", #the first neurons populating the developing cortical plate (preplate/subplate) 
"SOX2", "HOPX", "PEA15", "LGALS3BP", "MOXD1", #outer radial glia (oRG) 
"FEZF1", "NEUROD2", "NEUROD6", "TBR1", "PDE1A", #immature deep-layer projection neurons (DL PN) 
"SOX5", "LDB2", "CRYM", "TLE4", "BCL11B", "FEZF2", #deep-layer corticofugal projection neurons (CFuPN)
"UNC5A", "EPHA4", "BHLHE22", "PLXNA4", "SATB2", "CUX2", #callosal projection neurons (CPN) 
"GJA1", "S100B", "SPARC", "APOE", "AQP4", #glial cells with molecular features characteristic of astroglia
"OLIG1", "OLIG2", "PDGFRA", "EGFR", #oligodendrocyte precursors
"BIRC5", "DLX2", "DLX5", #cycling interneuron progenitors 
"GAD2", "DLX2", "DLX5", #immature interneurons 
"LAMP5", "LHX5", "GAD2", #subcortical interneurons
"PAX2", "CRABP1", "CRABP2", "TFAP2A", #subcortical interneurons, markers characteristic of caudal regions of the neural tube, i.e., hindbrain 
"FOXG1") #telencephalon / interneurons that appear after 4 months in culture
plot1 <- DoHeatmap(seurat_int_60, features = ct_markers) + guides(colour=FALSE)
plot2 <- DoHeatmap(seurat_int_120, features = ct_markers) + guides(colour=FALSE)
plot1 / plot2
```

```{r}
DefaultAssay(seurat_int_60) <- "RNA"
DefaultAssay(seurat_int_120) <- "RNA"
cl_markers_60_after <- FindAllMarkers(seurat_int_60, only.pos = TRUE, min.pct = 0.25, logfc.threshold = log(1.2))
top10_cl_markers_60_after <- cl_markers_60_after %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)
cl_markers_120_after <- FindAllMarkers(seurat_int_120, only.pos = TRUE, min.pct = 0.25, logfc.threshold = log(1.2))
top10_cl_markers_120_after <- cl_markers_120_after %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)
```

60 RRM2 in cluster 12
120 RRM2 in cluster 10

```{r}
DefaultAssay(seurat_int_60) <- "RNA"
DefaultAssay(seurat_int_120) <- "RNA"
FeaturePlot(seurat_int_60, features = c("S.Score", "G2M.Score"))
FeaturePlot(seurat_int_120, features = c("S.Score", "G2M.Score"))
```






```{r}
library(Seurat)
library(ggplot2)
library(patchwork)
```

```{r}
seurat_int_60[["RNA"]] <- as(seurat_int_60[["RNA"]], Class = "Assay")
class(seurat_int_60[["RNA"]])
seurat_int_120[["RNA"]] <- as(seurat_int_120[["RNA"]], Class = "Assay")
class(seurat_int_120[["RNA"]])
```

```{r}
DefaultAssay(seurat_int_60) <- "RNA"
DefaultAssay(seurat_int_120) <- "RNA"
seurat_int_60 <- CellCycleScoring(seurat_int_60,
 s.features = cc.genes$s.genes,
 g2m.features = cc.genes$g2m.genes,
 set.ident = TRUE)
seurat_int_120 <- CellCycleScoring(seurat_int_120,
 s.features = cc.genes$s.genes,
 g2m.features = cc.genes$g2m.genes,
 set.ident = TRUE)
```

```{r, fig.width=10, fig.height=10}
p1 <- VlnPlot(
  seurat_int_60,
  features = "S.Score",
  group.by = "seurat_clusters",
  pt.size = 0
) +
  geom_jitter(size = 0.2, alpha = 0.08, width = 0.3) +
  ggtitle("S.Score")

p2 <- VlnPlot(
  seurat_int_60,
  features = "G2M.Score",
  group.by = "seurat_clusters",
  pt.size = 0
) +
  geom_jitter(size = 0.2, alpha = 0.08, width = 0.3) +
  ggtitle("G2M.Score")

p1 + p2
```

```{r, fig.width=10, fig.height=10}
VlnPlot(seurat_int_60, features = c("S.Score", "G2M.Score"), group.by = "seurat_clusters", pt.size = 0)
FeaturePlot(seurat_int_60, features = c("S.Score", "G2M.Score"), ncol = 2)
DimPlot(seurat_int_60, reduction = "umap", group.by = "Phase")
table(seurat_int_60$seurat_clusters, seurat_int_60$Phase)
```

```{r}
DefaultAssay(seurat_int_60) <- "RNA"
DefaultAssay(seurat_int_120) <- "RNA"

```

```{r}
seurat_int_60[["RNA"]] <- as(seurat_int_60[["RNA"]], Class = "Assay")
class(seurat_int_60[["RNA"]])
seurat_int_120[["RNA"]] <- as(seurat_int_120[["RNA"]], Class = "Assay")
class(seurat_int_120[["RNA"]])
```

```{r}
DefaultAssay(seurat_int_60) <- "integrated"
DefaultAssay(seurat_int_120) <- "integrated"
seurat_int_60 <- ScaleData(seurat_int_60, vars.to.regress = c("S.Score", "G2M.Score"))
seurat_int_120 <- ScaleData(seurat_int_120, vars.to.regress = c("S.Score", "G2M.Score"))
```

```{r}
DefaultAssay(seurat_int_60) <- "integrated"
DefaultAssay(seurat_int_120) <- "integrated"
seurat_int_60 <- RunPCA(seurat_int_60, npcs = 50)
seurat_int_60 <- RunUMAP(seurat_int_60, dims = 1:30)
seurat_int_60 <- FindNeighbors(seurat_int_60, dims = 1:30) %>% FindClusters(resolution = 0.6)
seurat_int_120 <- RunPCA(seurat_int_120, npcs = 50)
seurat_int_120 <- RunUMAP(seurat_int_120, dims = 1:30)
seurat_int_120 <- FindNeighbors(seurat_int_120, dims = 1:30) %>% FindClusters(resolution = 0.6)
```



seurat_int_60_cc <- ScaleData(seurat_int_60_cc, vars.to.regress = c("S.Score", "G2M.Score"))
seurat_int_60_cc <- RunPCA(seurat_int_60_cc, npcs = 50)
seurat_int_60_cc <- RunUMAP(seurat_int_60_cc, dims = 1:30)
seurat_int_60_cc <- RunTSNE(seurat_int_60_cc, dims = 1:30)
seurat_int_60_cc <- FindNeighbors(seurat_int_60_cc, dims = 1:30) %>%
  FindClusters(resolution = 0.6)

DefaultAssay(seurat_int_60_cc) <- "RNA"

# Compare before vs after regression
plot1 <- DimPlot(seurat_int_60, reduction = "umap", label = TRUE)
plot2 <- DimPlot(seurat_int_60_cc, reduction = "umap", label = TRUE)
plot1 + plot2

# Phase distribution before regression
DimPlot(seurat_int_60_phase, reduction = "umap", group.by = "Phase")

# Optional marker checks after regression
FeaturePlot(seurat_int_60_cc, c("MKI67", "GLI3", "EOMES", "NEUROD6"), ncol = 4)

plot1 <- UMAPPlot(seurat_int_60_cc, group.by = "bio_sample")
plot2 <- UMAPPlot(seurat_int_60_cc, label = TRUE)
plot3 <- FeaturePlot(seurat_int_60_cc, c("SOX2", "STMN2", "HOPX", "SYT1"), ncol = 2, pt.size = 0.1)
((plot1 / plot2) | plot3) + plot_layout(widths = c(1, 2))





