---
title: "part1"
output: html_document
date: "2026-05-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(Seurat)
library(Matrix)
library(patchwork)
library(dplyr)
library(ggplot2)
library(pheatmap)

options(Seurat.object.assay.version = "v3")
```

```{r}
# Read the files
counts <- readMM("counts.mtx")
meta <- read.table("meta.tsv")
features <- read.table("features.tsv")[,1]
dimnames(counts) <- list(features, rownames(meta))
seurat <- CreateSeuratObject(counts = counts, meta.data = meta)

# How many cells and genes?
cat("Genes:", nrow(counts), "\n")
cat("Cells:", ncol(counts), "\n")

# Look at your metadata 
table(meta$sample, meta$condition)

``` 
Split by sample -> 4 sample/seurat
```{r}
seurat_list <- SplitObject(seurat, split.by = "sample")
```

Quality Control (QC) and filtering low-quality cells
```{r}
#Plotting "nFeature_RNA", "nCount_RNA", "percent.mt"
seurat_list <- lapply(seurat_list, function(x) {
  x[["percent.mt"]] <- PercentageFeatureSet(x, pattern = "^MT[-\\.]")
  x
})
lapply(seurat_list, function(x) {
  VlnPlot(x, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
})
lapply(seurat_list, function(x) {
  plot1 <- FeatureScatter(x, feature1 = "nCount_RNA", feature2 = "percent.mt") + ggtitle(unique(x$sample))
  plot2 <- FeatureScatter(x, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + ggtitle(unique(x$sample))
  plot1 + plot2
})

#Filter the cells by fixed threshold and find most 3000 differentially expressed genes
seurat_list <- lapply(seurat_list, function(x) {
  x <- subset(x, subset = nFeature_RNA > 500 & nFeature_RNA < 3500 & percent.mt < 10)
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, nfeatures = 3000)
  x <- ScaleData(x, vars.to.regress = c("nFeature_RNA", "percent.mt"))
  x
})
```

PCA over 4 samples
```{r}
seurat_list <- lapply(seurat_list, function(x) {
  x <- RunPCA(x, npcs = 50)
  x
})

#2 method to visualize data
#Elbow PCA
lapply(seurat_list, function(x) {
  ElbowPlot(x, ndims = ncol(Embeddings(x, "pca")))
})

#Heatmap PCA
lapply(seurat_list, function(x) {
  PCHeatmap(x, dims = 1:15, cells = 500, balanced = TRUE, ncol = 4)
})
message("PCA done")
```

Data Integration: YES or NO 
Is there a batch effects ? If cells of the same known type (e.g., progenitors or neurons/genes signature) from different replicates (only between WT or betw KO) fail to intermix on a UMAP or t-SNE plot, a batch effect is likely present.
```{r}
# Merge all samples without integration
seurat_merged <- merge(x = seurat_list$B1_17_1_day21, 
                       y = c(seurat_list$D2_22_3_day21, 
                             seurat_list$wt_2_day21, 
                             seurat_list$wt_4_day21),
                       add.cell.ids = c("B1", "D2", "wt2", "wt4")) %>%
  NormalizeData() %>%
  FindVariableFeatures(nfeatures = 3000) %>%
  ScaleData(vars.to.regress = c("nFeature_RNA", "percent.mt")) %>%
  RunPCA(npcs = 50) %>%
  RunUMAP(dims = 1:15)

# Plot by sample
DimPlot(seurat_merged, group.by = "sample")

# Plot by condition
DimPlot(seurat_merged, group.by = "condition")

# Subset for WT and plot
DimPlot(subset(seurat_merged, subset = condition == "WT"), 
                group.by = "sample", 
                pt.size = 0.5) + ggtitle("Wild Type Samples")

# Subset for KO and plot
DimPlot(subset(seurat_merged, subset = condition == "KO"), 
                group.by = "sample", 
                pt.size = 0.5) + ggtitle("KO Samples")

# Subset the object to only include Wild Type samples or KO
seurat_wt <- subset(seurat_merged, subset = condition == "WT")
seurat_ko <- subset(seurat_merged, subset = condition == "KO")

# --- WILD TYPE (WT) SAMPLES ---
FeaturePlot(seurat_wt, features = "POU5F1", split.by = "sample")
FeaturePlot(seurat_wt, features = "MIXL1", split.by = "sample")
FeaturePlot(seurat_wt, features = "EOMES", split.by = "sample")
FeaturePlot(seurat_wt, features = "HAND1", split.by = "sample")
FeaturePlot(seurat_wt, features = "MKI67", split.by = "sample")
FeaturePlot(seurat_wt, features = "ETV2", split.by = "sample")
FeaturePlot(seurat_wt, features = "TFPI2", split.by = "sample")
FeaturePlot(seurat_wt, features = "NRP2", split.by = "sample")
FeaturePlot(seurat_wt, features = "KDR", split.by = "sample")
FeaturePlot(seurat_wt, features = "CLDN5", split.by = "sample")
FeaturePlot(seurat_wt, features = "PECAM1", split.by = "sample")
FeaturePlot(seurat_wt, features = "GJA4", split.by = "sample")
FeaturePlot(seurat_wt, features = "GJA5", split.by = "sample")
FeaturePlot(seurat_wt, features = "CXCR4", split.by = "sample")
FeaturePlot(seurat_wt, features = "NR2F2", split.by = "sample")
FeaturePlot(seurat_wt, features = "APLNR", split.by = "sample")
FeaturePlot(seurat_wt, features = "PRCP", split.by = "sample")
FeaturePlot(seurat_wt, features = "DCN", split.by = "sample")
FeaturePlot(seurat_wt, features = "DLK1", split.by = "sample")
FeaturePlot(seurat_wt, features = "SULF1", split.by = "sample")
FeaturePlot(seurat_wt, features = "COL4A1", split.by = "sample")
FeaturePlot(seurat_wt, features = "ACTA2", split.by = "sample")
FeaturePlot(seurat_wt, features = "LUM", split.by = "sample")
FeaturePlot(seurat_wt, features = "PDGFRB", split.by = "sample")
FeaturePlot(seurat_wt, features = "CSPG4", split.by = "sample")

# --- KNOCKOUT (KO) SAMPLES ---
FeaturePlot(seurat_ko, features = "POU5F1", split.by = "sample")
FeaturePlot(seurat_ko, features = "MIXL1", split.by = "sample")
FeaturePlot(seurat_ko, features = "EOMES", split.by = "sample")
FeaturePlot(seurat_ko, features = "HAND1", split.by = "sample")
FeaturePlot(seurat_ko, features = "MKI67", split.by = "sample")
FeaturePlot(seurat_ko, features = "ETV2", split.by = "sample")
FeaturePlot(seurat_ko, features = "TFPI2", split.by = "sample")
FeaturePlot(seurat_ko, features = "NRP2", split.by = "sample")
FeaturePlot(seurat_ko, features = "KDR", split.by = "sample")
FeaturePlot(seurat_ko, features = "CLDN5", split.by = "sample")
FeaturePlot(seurat_ko, features = "PECAM1", split.by = "sample")
FeaturePlot(seurat_ko, features = "GJA4", split.by = "sample")
FeaturePlot(seurat_ko, features = "GJA5", split.by = "sample")
FeaturePlot(seurat_ko, features = "CXCR4", split.by = "sample")
FeaturePlot(seurat_ko, features = "NR2F2", split.by = "sample")
FeaturePlot(seurat_ko, features = "APLNR", split.by = "sample")
FeaturePlot(seurat_ko, features = "PRCP", split.by = "sample")
FeaturePlot(seurat_ko, features = "DCN", split.by = "sample")
FeaturePlot(seurat_ko, features = "DLK1", split.by = "sample")
FeaturePlot(seurat_ko, features = "SULF1", split.by = "sample")
FeaturePlot(seurat_ko, features = "COL4A1", split.by = "sample")
FeaturePlot(seurat_ko, features = "ACTA2", split.by = "sample")
FeaturePlot(seurat_ko, features = "LUM", split.by = "sample")
FeaturePlot(seurat_ko, features = "PDGFRB", split.by = "sample")
FeaturePlot(seurat_ko, features = "CSPG4", split.by = "sample")

saveRDS(seurat_merged, file="seurat_merged.rds")
message("Part1 Integration done - File saved")
```

Data Integration
```{r}
counts_int <- readMM("counts.mtx")
meta_int <- read.table("meta.tsv")
features_int <- read.table("features.tsv")[,1]
dimnames(counts_int) <- list(features_int, rownames(meta_int))
seurat_int <- CreateSeuratObject(counts = counts_int, meta.data = meta_int)
#Splitting dataset
seurat_list_int <- SplitObject(seurat_int, split.by = "sample") #KO vs WT
```

Filtering
```{r}
seurat_list_int <- lapply(seurat_list_int, function(x) {
  x[["percent.mt"]] <- PercentageFeatureSet(x, pattern = "^MT[-\\.]")
  x
})

seurat_list_int <- lapply(seurat_list_int, function(x) {
  x <- subset(x, subset = nFeature_RNA > 500 & nFeature_RNA < 3500 & percent.mt < 10)
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, nfeatures = 3000)
  x
})
```

FindIntegrationAnchors
```{r}
seurat_list_KO = seurat_list_int[c("B1_17_1_day21", "D2_22_3_day21")]
seurat_list_WT = seurat_list_int[c("wt_2_day21", "wt_4_day21")]
anchorsKO <- FindIntegrationAnchors(object.list = seurat_list_KO, dims = 1:15)
anchorsWT <- FindIntegrationAnchors(object.list = seurat_list_WT, dims = 1:15)
```

IntegrateData
```{r}
seurat_int_KO <- IntegrateData(anchorsKO, dims = 1:15)
seurat_int_WT <- IntegrateData(anchorsWT, dims = 1:15)
```

```{r}
seurat_int_KO <- ScaleData(seurat_int_KO)
seurat_int_KO <- RunPCA(seurat_int_KO, npcs = 15)
seurat_int_KO <- RunUMAP(seurat_int_KO, dims = 1:15)
seurat_int_KO <- RunTSNE(seurat_int_KO, dims = 1:15)
seurat_int_KO <- FindNeighbors(seurat_int_KO, dims = 1:15) %>% FindClusters(resolution = 0.6)
```

```{r}
seurat_int_WT <- ScaleData(seurat_int_WT)
seurat_int_WT <- RunPCA(seurat_int_WT, npcs = 15)
seurat_int_WT <- RunUMAP(seurat_int_WT, dims = 1:15)
seurat_int_WT <- RunTSNE(seurat_int_WT, dims = 1:15)
seurat_int_WT <- FindNeighbors(seurat_int_WT, dims = 1:15) %>% FindClusters(resolution = 0.6)
```

```{r}
DefaultAssay(seurat_int_WT) <- "integrated" #RNA for analysis (real counts) and integrated for visualization (corrected counts)
plot1 <- UMAPPlot(seurat_int_WT, group.by="sample")
plot2 <- UMAPPlot(seurat_int_WT, label = T)

# Use the real RNA expression values
DefaultAssay(seurat_int_WT) <- "RNA"
# FeaturePlot will now pull expression from the RNA assay,
# but the UMAP layout is the one computed from the integrated assay
plot3 <- FeaturePlot(seurat_int_WT, c("PECAM1","CD34","CLDN5"), ncol = 2, pt.size = 0.1)
((plot1 / plot2) | plot3) + plot_layout(width = c(1,2))
```

```{r}
DefaultAssay(seurat_int_KO) <- "integrated"
plot1 <- UMAPPlot(seurat_int_KO, group.by="sample")
plot2 <- UMAPPlot(seurat_int_KO, label = T)

# Use the real RNA expression values
DefaultAssay(seurat_int_WT) <- "RNA"
# FeaturePlot will now pull expression from the RNA assay,
# but the UMAP layout is the one computed from the integrated assay
plot3 <- FeaturePlot(seurat_int_KO, c("PECAM1","CD34","CLDN5"), ncol = 2, pt.size = 0.1)
((plot1 / plot2) | plot3) + plot_layout(width = c(1,2))

message("Part2 Integration done")
```


Cluster Annotation based on manually curated genes (Supplementary Table 1 and others added)
```{r}
# choose Seurat Object to use
seurat_obj <- seurat_int_KO   # seurat_int_WT or seurat_int_KO

# List of your genes -> from paper and manually added
genes_of_interest <- c(
  "POU5F1", "THRB", "MKI67", "HAND2", "BAMBI", "LUM", "DCN", "DLK1", 
  "ETV2", "MAML1", "PITX1", "SPRY1", "TRPS1", "SNAI2", "PDGFRB", "CSPG4", 
  "ANGPT2", "MATN4", "MIXL1", "DLL3", "SULF1", "COL4A1", "LOX", "KRT19", 
  "WNT5A", "NID2", "POSTN", "IGFBP3", "COL5A1", "HES2", "HRH1", "SMAD3", 
  "RGCC", "PLAUR", "IGFBP5", "NR3C1", "GATA6", "SGK1", "DKK1", "CRABP1", 
  "HEY1", "DLL4", "VWF", "CLDN5", "CXCR4", "GJA4", "GJA5", "SEMA3G", 
  "MCAM", "WFDC2", "DLL1", "ARL4A", "ZFPM2", "HES1", "PRDM1", "FOXO1", 
  "ESM1", "MECOM", "PECAM1", "ADM", "SOX18", "ERG", "PDGFB", "MMRN1", 
  "CDH5", "DUSP23", "CFD", "SOX7", "KDR", "ROBO4", "PLVAP", "FLI1", 
  "HHEX", "GATA2", "ELK3", "LEF1", "CPE", "EPCAM", "HES4", "PDGFA", 
  "RGS5", "PGF", "HES5", "MYC", "RBPJ", "EP300", "SPINT2", "ZEB2", 
  "EBF1", "NR2F2", "NRP2", "APLNR", "ZEB1", "ADAMTS18", "NTS", "NEFH", 
  "LY6H", "RGS16", "HAND1", "KCNE3", "TFPI2", "MESP1", "MESP2", "PTPRC", 
  "ALDH1A1", "CTGF", "FOXQ1", "PAX3", "AQP4", "ANKRD1", "TAGLN", "ACTA2"
)

#Keep only genes that exist in your Seurat object and get average expression per cluster 
genes_present <- intersect(genes_of_interest, rownames(seurat_obj))
avg_expr <- AverageExpression(seurat_obj, features = genes_present, group.by = "seurat_clusters", slot = "data")
avg_matrix <- as.matrix(avg_expr$RNA)
z_score <- t(scale(t(avg_matrix)))
colnames(z_score) <- gsub("g", "", colnames(z_score)) #"g0" becomes "0", for style
# Manually ordered to obtain a red diagonal
my_cluster_order <- c("0", "2", "9", "3", "8", "5",   
                      "1", "7", "4", "10", "6", "11") 
my_order <- intersect(my_cluster_order, colnames(z_score))
z_score_ordered <- z_score[, my_order]
var_sub <- apply(z_score_ordered, 1, var, na.rm = TRUE)
z_score_ordered <- z_score_ordered[var_sub > 0 & !is.na(var_sub), ]

# heatmap
h <- pheatmap(z_score_ordered,
         scale = "none",
         cluster_rows = TRUE,      # Keeps your custom order
         cluster_cols = FALSE, 
         treeheight_row = 0,        # Deletes the row dendrogram
         treeheight_col = 0,        # Deletes the column dendrogram (if any)
         fontsize_row = 5,          # Prevents name overlap
         height = 2,
         angle_col = 0,             # Keeps cluster names horizontal
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "Z-scored gene expression per cluster ")

# Save pdf
ggsave("heatmap.pdf", plot = h, width = 12, height = 8, dpi=600,  bg = "white")

# Printing cluster's marker gene to ask LLMs which tissue corresponds
genes_main <- apply(z_score_ordered, 2, function(x) {
names(x[x > 2]) })  #Filter Z-score > 2
genes_minor <- apply(z_score_ordered, 2, function(x) {
names(x[x > 1 & x < 2]) }) #Filter z-score > 1 & z-score < 2

# Print main and minor genes per cluster -> ASK LLMs to get cell type and cell state (if inferable)
print(genes_main)
print(genes_minor)

```

Cell Annotation Assignment
```{r}
# --------------- WT cluster identity --------------- #
Idents(seurat_int_WT) <- "seurat_clusters"
new_cluster_ids <- c(
  "0" = "Fibroblast-like Mural",
  "1" = "Mature Mural",
  "2" = "Epicardial Progenitors",
  "3" = "Epicardial-like",
  "4" = "Neural Crest",
  "5" = "Mesoderm Progenitors",
  "6" = "Venous/Capillary ECs",
  "7" = "Capillary/Tip-like",
  "8" = "Mural Progenitors",
  "9" = "Proliferating Pluripotent",
  "10" = "Neuromesoderm-like",
  "11" = "Pan-Endothelial"
)
seurat_int_WT <- RenameIdents(seurat_int_WT, new_cluster_ids)
seurat_int_WT$cell_type <- Idents(seurat_int_WT)

# --------------- KO cluster identity --------------- #
Idents(seurat_int_KO) <- "seurat_clusters"
new_cluster_ids <- c(
  "0"  = "Epicardial-like",
  "1"  = "Fibroblast-like Mural",
  "2"  = "Unknown",
  "3"  = "Mural Progenitors",
  "4"  = "Mixed Off-Target Progenitors (NM/Epi/NC)",
  "5"  = "Neural Crest",
  "6"  = "Mesoderm Progenitors",
  "7"  = "Proliferating Epicardial Progenitors",
  "8"  = "Contractile pericytes",
  "9"  = "Proliferating Arterial Endothelial",
  "10" = "Pan-Endothelial"
)
seurat_int_KO <- RenameIdents(seurat_int_KO, new_cluster_ids)
seurat_int_KO$cell_type <- Idents(seurat_int_KO)

# plots with annotation
DimPlot(seurat_int_WT, reduction = "umap", label = FALSE) + ggtitle("WT Day 21 Clusters")
DimPlot(seurat_int_KO, reduction = "umap", label = FALSE) + ggtitle("MECOM KO Day 21 Clusters")
```

# Report from Gemini Pro 3.1 about its cell annotation reasoning

WT (seurat_int_WT)
Cluster 0: Fibroblast-like Mural Cells** (Core markers: *DCN, IGFBP5, TRPS1*).
Cluster 1: Mature Mural Cells / ECM** (Matrix heavy: *COL5A1, POSTN, COL4A1*).
Cluster 2: Epicardial / Lateral Plate Mesoderm Progenitors** (Core markers: *HAND1, KCNE3*).
Cluster 3: Epicardial-like / Mesothelial** (Core markers: *EPCAM, ADAMTS18, DLK1*).
Cluster 4: Neural Crest** (Core markers: *PAX3, FOXQ1, AQP4*).
Cluster 5: Early Mesoderm Progenitor** (Core markers: *MIXL1, HAND1*).
Cluster 6: Venous / Capillary Endothelial Cells** (Core EC markers mixed with *NRP2, ARL4A, NR2F2*).
Cluster 7: Capillary / Tip-like Cells** (Confidence: Low. *RGCC* indicates capillary, but weak overall signature).
Cluster 8: Mural Progenitors / Pericytes** (Core markers: *PDGFRB, CSPG4, RGS5*).
Cluster 9: Proliferating Pluripotent Cells** (Core markers: *POU5F1, MKI67*).
Cluster 10: Neuromesoderm-like / Somitic** (Core markers: *MESP1, MESP2*).
Cluster 11: Pan-Endothelial (Arterial dominant)** (Massive definitive EC signature: *PECAM1, CDH5, DLL4, GJA5*).

KO (seurat_int_KO)
Cluster 0: Epicardial-like / Mesenchymal Progenitor
Main Markers: KCNE3, IGFBP3.
Minor Markers: DLK1, KRT19, LUM, COL5A1.
Cluster 1: Fibroblast-like Mural Cells / ECM (Matrix-producing Mural Cells)
Main Markers: WNT5A
Minor Markers: POSTN, COL4A1, COL5A1, SULF1, LOX.
Cluster 2: Unknown
Main Markers: None.
Cluster 3: Early Mural Progenitors
Main Markers: ADAMTS18
Minor Markers: PDGFRB, DCN, ETV2.
Cluster 4: Mixed Off-Target Progenitors -  Neuromesoderm (MESP1/2, DLL3), Epithelial/Endoderm (EPCAM), and Neural Crest (CRABP1)
Main Markers: EPCAM, MIXL1, MESP1, MESP2, DLL3, CRABP1, PTPRC.
Cluster 5: Neural Crest / Mesenchymal
Main Markers: PAX3
Minor Markers: TRPS1, HAND2, IGFBP5, CSPG4.
Cluster 6: Early Mesoderm Progenitors
Main Markers: None
Minor Markers: MIXL1, HAND1, ETV2.
Cluster 7: Proliferating Epicardial Progenitors
Main Markers: None
Minor Markers: PITX1, HAND1, MKI67, MATN4.
Cluster 8: Contractile / Mature Mural Cells (Pericytes/SMCs)
Main Markers: TAGLN, ACTA2 (->Mature and conctractile state), ANKRD1, COL4A1, PDGFA.
Minor Markers: POSTN, MCAM, SNAI2.
Cluster 9: Proliferating Arterial Endothelial Cells
Main Markers: MKI67.
Minor: GJA5 (arterial marker).
Cluster 10: Pan-Endothelial Cells
Main Markers: PECAM1, CDH5, KDR, FLI1, ERG (Core EC), DLL4, GJA5, CXCR4 (Arterial), VWF, NRP2 (Venous), PRDM1, PLVAP (Capillary)

Save for Pyhton
```{r}
# ------------------- export to R ------------------- #
saveRDS(seurat_int_KO, file="seurat_int_KO.rds")
saveRDS(seurat_int_WT, file="seurat_int_WT.rds")
# --------------------------------------------------- #


# ------------------- export to python ------------------- #
library(SingleCellExperiment)
library(zellkonverter)

DefaultAssay(seurat_int_KO) <- "RNA"
seurat_int_KO <- JoinLayers(seurat_int_KO)
sce_ko <- as.SingleCellExperiment(seurat_int_KO)
reducedDim(sce_ko, "UMAP") <- Embeddings(seurat_int_KO, "umap")
reducedDim(sce_ko, "PCA") <- Embeddings(seurat_int_KO, "pca")
writeH5AD(sce_ko, "seurat_int_KO.h5ad")

DefaultAssay(seurat_int_WT) <- "RNA"
seurat_int_WT <- JoinLayers(seurat_int_WT)
sce_wt <- as.SingleCellExperiment(seurat_int_WT)
reducedDim(sce_wt, "UMAP") <- Embeddings(seurat_int_WT, "umap")
reducedDim(sce_wt, "PCA") <- Embeddings(seurat_int_WT, "pca")
writeH5AD(sce_wt, "seurat_int_WT.h5ad")
# -------------------------------------------------------- #
```












