---
title: "part1"
output: html_document
date: "`r Sys.Date()`"
---


```{r setup, include=FALSE}
# This sets the directory for ALL subsequent chunks
knitr::opts_knit$set(root.dir = '/Users/marco/Desktop/SingleCells/part1')
```

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(Matrix)
  library(patchwork)
  library(dplyr)
  library(ggplot2)
  library(pheatmap)
  library(SingleCellExperiment)
  library(zellkonverter)
  #library(simspec)
})

options(Seurat.object.assay.version = "v3")
```


```{r}
# Read the files
counts <- readMM("counts.mtx")
meta <- read.table("meta.tsv")
features <- read.table("features.tsv")[,1]
dimnames(counts) <- list(features, rownames(meta))
seurat <- CreateSeuratObject(counts = counts, meta.data = meta)

# How many cells and genes?
cat("Genes:", nrow(counts), "\n")
cat("Cells:", ncol(counts), "\n")

# Look at your metadata 
table(meta$sample, meta$condition)
``` 
Split by sample -> 4 sample/seurat
```{r}
seurat_list <- SplitObject(seurat, split.by = "sample")
```

Quality Control (QC) and filtering low-quality cells
```{r}
#Plotting "nFeature_RNA", "nCount_RNA", "percent.mt"
seurat_list <- lapply(seurat_list, function(x) {
  x[["percent.mt"]] <- PercentageFeatureSet(x, pattern = "^MT[-\\.]")
  x
})
lapply(seurat_list, function(x) {
  VlnPlot(x, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
})
lapply(seurat_list, function(x) {
  plot1 <- FeatureScatter(x, feature1 = "nCount_RNA", feature2 = "percent.mt") + ggtitle(unique(x$sample))
  plot2 <- FeatureScatter(x, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + ggtitle(unique(x$sample))
  plot1 + plot2
})

#Filter the cells by fixed threshold and find most 3000 differentially expressed genes
seurat_list <- lapply(seurat_list, function(x) {
  x <- subset(x, subset = nFeature_RNA > 500 & nFeature_RNA < 6000 & percent.mt < 10)
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, nfeatures = 3000)
  x <- ScaleData(x, vars.to.regress = c("nFeature_RNA", "percent.mt"))
  x
})
```

PCA over 4 samples
```{r}
seurat_list <- lapply(seurat_list, function(x) {
  x <- RunPCA(x, npcs = 50)
  x
})

#2 method to visualize data
#Elbow PCA
lapply(seurat_list, function(x) {
  ElbowPlot(x, ndims = ncol(Embeddings(x, "pca")))
})

#Heatmap PCA
lapply(seurat_list, function(x) {
  PCHeatmap(x, dims = 1:15, cells = 500, balanced = TRUE, ncol = 4)
})
message("PCA done")
```

Data Integration: YES or NO 
Is there a batch effects ? If cells of the same known type (e.g., progenitors or neurons/genes signature) from different replicates (only between WT or betw KO) fail to intermix on a UMAP or t-SNE plot, a batch effect is likely present.
```{r}
# Merge all samples without integration
seurat_merged <- merge(x = seurat_list$B1_17_1_day21, 
                       y = c(seurat_list$D2_22_3_day21, 
                             seurat_list$wt_2_day21, 
                             seurat_list$wt_4_day21),
                       add.cell.ids = c("B1", "D2", "wt2", "wt4")) %>%
  NormalizeData() %>%
  FindVariableFeatures(nfeatures = 3000) %>%
  ScaleData(vars.to.regress = c("nFeature_RNA", "percent.mt")) %>%
  RunPCA(npcs = 50) %>%
  RunUMAP(dims = 1:15)

# Plot by sample
DimPlot(seurat_merged, group.by = "sample")

# Plot by condition
DimPlot(seurat_merged, group.by = "condition")

# Subset for WT and plot
DimPlot(subset(seurat_merged, subset = condition == "WT"), 
                group.by = "sample", 
                pt.size = 0.5) + ggtitle("Wild Type Samples")

# Subset for KO and plot
DimPlot(subset(seurat_merged, subset = condition == "KO"), 
                group.by = "sample", 
                pt.size = 0.5) + ggtitle("KO Samples")

# Subset the object to only include Wild Type samples or KO
seurat_wt <- subset(seurat_merged, subset = condition == "WT")
seurat_ko <- subset(seurat_merged, subset = condition == "KO")

# --- WILD TYPE (WT) SAMPLES ---
FeaturePlot(seurat_wt, features = "MKI67", split.by = "sample")
FeaturePlot(seurat_wt, features = "TOP2A", split.by = "sample")
FeaturePlot(seurat_wt, features = "CDK1", split.by = "sample")
FeaturePlot(seurat_wt, features = "CCNB1", split.by = "sample")
FeaturePlot(seurat_wt, features = "PCNA", split.by = "sample")
FeaturePlot(seurat_wt, features = "RRM2", split.by = "sample")

# --- KNOCKOUT (KO) SAMPLES ---
FeaturePlot(seurat_ko, features = "MKI67", split.by = "sample")
FeaturePlot(seurat_ko, features = "TOP2A", split.by = "sample")
FeaturePlot(seurat_ko, features = "CDK1", split.by = "sample")
FeaturePlot(seurat_ko, features = "CCNB1", split.by = "sample")
FeaturePlot(seurat_ko, features = "PCNA", split.by = "sample")
FeaturePlot(seurat_ko, features = "RRM2", split.by = "sample")

saveRDS(seurat_merged, file="seurat_merged.rds")
message("Part1 Integration done - File saved")
```

```{r}

# --- WILD TYPE (WT) SAMPLES ---
FeaturePlot(seurat_wt, features = "MKI67", split.by = "sample")
FeaturePlot(seurat_wt, features = "TOP2A", split.by = "sample")
FeaturePlot(seurat_wt, features = "CDK1", split.by = "sample")
FeaturePlot(seurat_wt, features = "CCNB1", split.by = "sample")
FeaturePlot(seurat_wt, features = "PCNA", split.by = "sample")
FeaturePlot(seurat_wt, features = "RRM2", split.by = "sample")

# --- KNOCKOUT (KO) SAMPLES ---
FeaturePlot(seurat_ko, features = "MKI67", split.by = "sample")
FeaturePlot(seurat_ko, features = "TOP2A", split.by = "sample")
FeaturePlot(seurat_ko, features = "CDK1", split.by = "sample")
FeaturePlot(seurat_ko, features = "CCNB1", split.by = "sample")
FeaturePlot(seurat_ko, features = "PCNA", split.by = "sample")
FeaturePlot(seurat_ko, features = "RRM2", split.by = "sample")

```


Data Integration
```{r}
counts_int <- readMM("counts.mtx")
meta_int <- read.table("meta.tsv")
features_int <- read.table("features.tsv")[,1]
dimnames(counts_int) <- list(features_int, rownames(meta_int))
seurat_int <- CreateSeuratObject(counts = counts_int, meta.data = meta_int)
#Splitting dataset
seurat_list_int <- SplitObject(seurat_int, split.by = "sample") #KO vs WT
```

```{r}
seurat_list_int <- lapply(seurat_list_int, function(x) {
  x[["percent.mt"]] <- PercentageFeatureSet(x, pattern = "^MT[-\\.]")
  x
})
```

```{r}
seurat_list_int <- lapply(seurat_list_int, function(x) {
  x <- subset(x, subset = nFeature_RNA > 500 & nFeature_RNA < 6000 & percent.mt < 10)
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, nfeatures = 3000)
  x <- CellCycleScoring(x, s.features = cc.genes.updated.2019$s.genes,
                   g2m.features = cc.genes.updated.2019$g2m.genes)
  VariableFeatures(x) <- setdiff(VariableFeatures(x),
                                    c(unlist(cc.genes.updated.2019),
                                    grep("^MT-", rownames(x), value=T)))
  x
})
```

FindIntegrationAnchors
```{r}
seurat_list_KO = seurat_list_int[c("B1_17_1_day21", "D2_22_3_day21")]
seurat_list_WT = seurat_list_int[c("wt_2_day21", "wt_4_day21")]
anchorsKO <- FindIntegrationAnchors(object.list = seurat_list_KO, dims = 1:15)
anchorsWT <- FindIntegrationAnchors(object.list = seurat_list_WT, dims = 1:15)
```

IntegrateData
```{r}
seurat_int_KO <- IntegrateData(anchorsKO, dims = 1:15)
seurat_int_WT <- IntegrateData(anchorsWT, dims = 1:15)
```

```{r}
seurat_int_KO <- ScaleData(seurat_int_KO, vars.to.regress = c("G2M.Score","S.Score")) 
seurat_int_KO <- RunPCA(seurat_int_KO, npcs = 15)
seurat_int_KO <- RunUMAP(seurat_int_KO, dims = 1:15)
seurat_int_KO <- RunTSNE(seurat_int_KO, dims = 1:15)
seurat_int_KO <- FindNeighbors(seurat_int_KO, dims = 1:15) %>% FindClusters(resolution = 0.6)
```

```{r}
seurat_int_WT <- ScaleData(seurat_int_WT, vars.to.regress = c("G2M.Score","S.Score"))
seurat_int_WT <- RunPCA(seurat_int_WT, npcs = 15)
seurat_int_WT <- RunUMAP(seurat_int_WT, dims = 1:15)
seurat_int_WT <- RunTSNE(seurat_int_WT, dims = 1:15)
seurat_int_WT <- FindNeighbors(seurat_int_WT, dims = 1:15) %>% FindClusters(resolution = 0.6)
```

```{r}
DefaultAssay(seurat_int_WT) <- "integrated" #RNA for analysis (real counts) and integrated for visualization (corrected counts)
plot1 <- UMAPPlot(seurat_int_WT, group.by="sample")
plot2 <- UMAPPlot(seurat_int_WT, label = T)

# Use the real RNA expression values
DefaultAssay(seurat_int_WT) <- "RNA"
# FeaturePlot will now pull expression from the RNA assay,
# but the UMAP layout is the one computed from the integrated assay
plot3 <- FeaturePlot(seurat_int_WT, c("PECAM1","CD34","CLDN5"), ncol = 2, pt.size = 0.1)
((plot1 / plot2) | plot3) + plot_layout(width = c(1,2))
```

```{r}
DefaultAssay(seurat_int_KO) <- "integrated"
plot1 <- UMAPPlot(seurat_int_KO, group.by="sample")
plot2 <- UMAPPlot(seurat_int_KO, label = T)

# Use the real RNA expression values
DefaultAssay(seurat_int_WT) <- "RNA"
# FeaturePlot will now pull expression from the RNA assay,
# but the UMAP layout is the one computed from the integrated assay
plot3 <- FeaturePlot(seurat_int_KO, c("PECAM1","CD34","CLDN5"), ncol = 2, pt.size = 0.1)
((plot1 / plot2) | plot3) + plot_layout(width = c(1,2))

message("Part2 Integration done")
```

Cluster Annotation based on manually curated genes (Supplementary Table 1 and others added)
```{r}
# choose Seurat Object to use
seurat_obj <- seurat_int_WT   # seurat_int_WT or seurat_int_KO

# List of your genes -> from paper and manually added
genes_of_interest <- c(
  "POU5F1", "THRB", "MKI67","MKI67", "TOP2A", "CDK1", "CCNB1", "PCNA", "RRM2", "HAND2", "BAMBI", "LUM", "DCN", "DLK1", 
  "ETV2", "MAML1", "PITX1", "SPRY1", "TRPS1", "SNAI2", "PDGFRB", "CSPG4", 
  "ANGPT2", "MATN4", "MIXL1", "DLL3", "SULF1", "COL4A1", "LOX", "KRT19", 
  "WNT5A", "NID2", "POSTN", "IGFBP3", "COL5A1", "HES2", "HRH1", "SMAD3", 
  "RGCC", "PLAUR", "IGFBP5", "NR3C1", "GATA6", "SGK1", "DKK1", "CRABP1", 
  "HEY1", "DLL4", "VWF", "CLDN5", "CXCR4", "GJA4", "GJA5", "SEMA3G", 
  "MCAM", "WFDC2", "DLL1", "ARL4A", "ZFPM2", "HES1", "PRDM1", "FOXO1", 
  "ESM1", "MECOM", "PECAM1", "ADM", "SOX18", "ERG", "PDGFB", "MMRN1", 
  "CDH5", "DUSP23", "CFD", "SOX7", "KDR", "ROBO4", "PLVAP", "FLI1", 
  "HHEX", "GATA2", "ELK3", "LEF1", "CPE", "EPCAM", "HES4", "PDGFA", 
  "RGS5", "PGF", "HES5", "MYC", "RBPJ", "EP300", "SPINT2", "ZEB2", 
  "EBF1", "NR2F2", "NRP2", "APLNR", "ZEB1", "ADAMTS18", "NTS", "NEFH", 
  "LY6H", "RGS16", "HAND1", "KCNE3", "TFPI2", "MESP1", "MESP2", "PTPRC", 
  "ALDH1A1", "CTGF", "FOXQ1", "PAX3", "AQP4", "ANKRD1", "TAGLN", "ACTA2"
)

#Keep only genes that exist in your Seurat object and get average expression per cluster 
genes_present <- intersect(genes_of_interest, rownames(seurat_obj))
avg_expr <- AverageExpression(seurat_obj, features = genes_present, group.by = "seurat_clusters", slot = "data")
avg_matrix <- as.matrix(avg_expr$RNA)
z_score <- t(scale(t(avg_matrix)))
colnames(z_score) <- gsub("g", "", colnames(z_score)) #"g0" becomes "0", for style
# Manually ordered to obtain a red diagonal
my_cluster_order <- c("7", "10", "3", "9", "4", "6", "5", "1", "2", "8", "0", "11") 
my_order <- intersect(my_cluster_order, colnames(z_score))
z_score_ordered <- z_score[, my_order]
var_sub <- apply(z_score_ordered, 1, var, na.rm = TRUE)
z_score_ordered <- z_score_ordered[var_sub > 0 & !is.na(var_sub), ]

# heatmap
h <- pheatmap(z_score_ordered,
         scale = "none",
         cluster_rows = TRUE,       # Keeps your custom order
         cluster_cols = FALSE, 
         treeheight_row = 0,        # Deletes the row dendrogram
         treeheight_col = 0,        # Deletes the column dendrogram (if any)
         fontsize_row = 5,          # Prevents name overlap
         height = 2,
         angle_col = 0,             # Keeps cluster names horizontal
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "Z-scored gene expression per cluster ")

# Save pdf
ggsave("heatmap.pdf", plot = h, width = 12, height = 8, dpi=600,  bg = "white")

# Printing cluster's marker gene to ask LLMs which tissue corresponds
genes_main <- apply(z_score_ordered, 2, function(x) {
names(x[x > 2]) })  #Filter Z-score > 2
genes_minor <- apply(z_score_ordered, 2, function(x) {
names(x[x > 1 & x < 2]) }) #Filter z-score > 1 & z-score < 2

# Print main and minor genes per cluster -> ASK LLMs to get cell type and cell state (if inferable)
print(genes_main)
print(genes_minor)

```

Cell Annotation Assignment
```{r, include=FALSE, eval=FALSE}
# --------------- WT cluster identity --------------- #
Idents(seurat_int_WT) <- "seurat_clusters"
new_cluster_ids <- c(
  "0" = "Fibroblast-like Mural",
  "1" = "Mature Mural",
  "2" = "Epicardial Progenitors",
  "3" = "Epicardial-like",
  "4" = "Neural Crest",
  "5" = "Mesoderm Progenitors",
  "6" = "Venous/Capillary ECs",
  "7" = "Capillary/Tip-like",
  "8" = "Mural Progenitors",
  "9" = "Proliferating Pluripotent",
  "10" = "Neuromesoderm-like",
  "11" = "Pan-Endothelial"
)
seurat_int_WT <- RenameIdents(seurat_int_WT, new_cluster_ids)
seurat_int_WT$cell_type <- Idents(seurat_int_WT)

# --------------- KO cluster identity --------------- #
Idents(seurat_int_KO) <- "seurat_clusters"
new_cluster_ids <- c(
  "0"  = "Epicardial-like",
  "1"  = "Fibroblast-like Mural",
  "2"  = "Unknown",
  "3"  = "Mural Progenitors",
  "4"  = "Mixed Off-Target Progenitors (NM/Epi/NC)",
  "5"  = "Neural Crest",
  "6"  = "Mesoderm Progenitors",
  "7"  = "Proliferating Epicardial Progenitors",
  "8"  = "Contractile pericytes",
  "9"  = "Proliferating Arterial Endothelial",
  "10" = "Pan-Endothelial"
)
seurat_int_KO <- RenameIdents(seurat_int_KO, new_cluster_ids)
seurat_int_KO$cell_type <- Idents(seurat_int_KO)

# plots with annotation
DimPlot(seurat_int_WT, reduction = "umap", label = FALSE) + ggtitle("WT Day 21 Clusters")
DimPlot(seurat_int_KO, reduction = "umap", label = FALSE) + ggtitle("MECOM KO Day 21 Clusters")
```

# Report from Gemini Pro 3.1 about its cell annotation reasoning

WT (seurat_int_WT)
Cluster 0: Fibroblast-like Mural Cells** (Core markers: *DCN, IGFBP5, TRPS1*).
Cluster 1: Mature Mural Cells / ECM** (Matrix heavy: *COL5A1, POSTN, COL4A1*).
Cluster 2: Epicardial / Lateral Plate Mesoderm Progenitors** (Core markers: *HAND1, KCNE3*).
Cluster 3: Epicardial-like / Mesothelial** (Core markers: *EPCAM, ADAMTS18, DLK1*).
Cluster 4: Neural Crest** (Core markers: *PAX3, FOXQ1, AQP4*).
Cluster 5: Early Mesoderm Progenitor** (Core markers: *MIXL1, HAND1*).
Cluster 6: Venous / Capillary Endothelial Cells** (Core EC markers mixed with *NRP2, ARL4A, NR2F2*).
Cluster 7: Capillary / Tip-like Cells** (Confidence: Low. *RGCC* indicates capillary, but weak overall signature).
Cluster 8: Mural Progenitors / Pericytes** (Core markers: *PDGFRB, CSPG4, RGS5*).
Cluster 9: Proliferating Pluripotent Cells** (Core markers: *POU5F1, MKI67*).
Cluster 10: Neuromesoderm-like / Somitic** (Core markers: *MESP1, MESP2*).
Cluster 11: Pan-Endothelial (Arterial dominant)** (Massive definitive EC signature: *PECAM1, CDH5, DLL4, GJA5*).

KO (seurat_int_KO)
Cluster 0: Epicardial-like / Mesenchymal Progenitor
Main Markers: KCNE3, IGFBP3.
Minor Markers: DLK1, KRT19, LUM, COL5A1.
Cluster 1: Fibroblast-like Mural Cells / ECM (Matrix-producing Mural Cells)
Main Markers: WNT5A
Minor Markers: POSTN, COL4A1, COL5A1, SULF1, LOX.
Cluster 2: Unknown
Main Markers: None.
Cluster 3: Early Mural Progenitors
Main Markers: ADAMTS18
Minor Markers: PDGFRB, DCN, ETV2.
Cluster 4: Mixed Off-Target Progenitors -  Neuromesoderm (MESP1/2, DLL3), Epithelial/Endoderm (EPCAM), and Neural Crest (CRABP1)
Main Markers: EPCAM, MIXL1, MESP1, MESP2, DLL3, CRABP1, PTPRC.
Cluster 5: Neural Crest / Mesenchymal
Main Markers: PAX3
Minor Markers: TRPS1, HAND2, IGFBP5, CSPG4.
Cluster 6: Early Mesoderm Progenitors
Main Markers: None
Minor Markers: MIXL1, HAND1, ETV2.
Cluster 7: Proliferating Epicardial Progenitors
Main Markers: None
Minor Markers: PITX1, HAND1, MKI67, MATN4.
Cluster 8: Contractile / Mature Mural Cells (Pericytes/SMCs)
Main Markers: TAGLN, ACTA2 (->Mature and conctractile state), ANKRD1, COL4A1, PDGFA.
Minor Markers: POSTN, MCAM, SNAI2.
Cluster 9: Proliferating Arterial Endothelial Cells
Main Markers: MKI67.
Minor: GJA5 (arterial marker).
Cluster 10: Pan-Endothelial Cells
Main Markers: PECAM1, CDH5, KDR, FLI1, ERG (Core EC), DLL4, GJA5, CXCR4 (Arterial), VWF, NRP2 (Venous), PRDM1, PLVAP (Capillary)

Save for Pyhton
```{r, include=FALSE, eval=FALSE}
# ------------------- export to R ------------------- #
saveRDS(seurat_int_KO, file="seurat_int_KO.rds")
saveRDS(seurat_int_WT, file="seurat_int_WT.rds")
# --------------------------------------------------- #

# ------------------- export to python ------------------- #
DefaultAssay(seurat_int_KO) <- "RNA"
seurat_int_KO <- JoinLayers(seurat_int_KO)
sce_ko <- as.SingleCellExperiment(seurat_int_KO)
reducedDim(sce_ko, "UMAP") <- Embeddings(seurat_int_KO, "umap")
reducedDim(sce_ko, "PCA") <- Embeddings(seurat_int_KO, "pca")
writeH5AD(sce_ko, "seurat_int_KO.h5ad")

DefaultAssay(seurat_int_WT) <- "RNA"
seurat_int_WT <- JoinLayers(seurat_int_WT)
sce_wt <- as.SingleCellExperiment(seurat_int_WT)
reducedDim(sce_wt, "UMAP") <- Embeddings(seurat_int_WT, "umap")
reducedDim(sce_wt, "PCA") <- Embeddings(seurat_int_WT, "pca")
writeH5AD(sce_wt, "seurat_int_WT.h5ad")
# -------------------------------------------------------- #
```

Whole-Integration
```{r}
# From seurat_list_int already filtered
anchors <- FindIntegrationAnchors(object.list = seurat_list_int, dims = 1:15)
seurat_int_all <- IntegrateData(anchors, dims = 1:15)
seurat_int_all <- ScaleData(seurat_int_all, vars.to.regress = c("G2M.Score","S.Score"))
seurat_int_all <- RunPCA(seurat_int_all)
seurat_int_all <- RunUMAP(seurat_int_all, dims = 1:15)
seurat_int_all <- FindNeighbors(seurat_int_all, dims = 1:15)
seurat_int_all <- FindClusters(seurat_int_all, resolution = 0.6)
#saveRDS(seurat_int_all, file="seurat_int_all.rds") 

# Plot whole-integration
DimPlot(seurat_integrated, reduction = "umap", label = TRUE)
DimPlot(seurat_integrated, reduction = "umap", group.by = "condition")
DimPlot(seurat_integrated, reduction = "umap", split.by = "condition")

# Histogram of WT/KO cell proportion per cluster
counts <- table(seurat_integrated$seurat_clusters, seurat_integrated$condition)
proportions <- as.data.frame(prop.table(counts, margin = 1))
colnames(proportions) <- c("Cluster", "Condition", "Proportion")
ggplot(proportions, aes(x = Cluster, y = Proportion, fill = Condition)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  labs(y = "Fraction of Cells", title = "Proportion of WT vs KO per Cluster") +
  scale_y_continuous(labels = scales::percent)
```

Cluster Annotation based on manually curated genes (Supplementary Table 1 and others added) - on whole Integration
```{r}
# choose Seurat Object to use
seurat_obj <- cluster5_sub    # seurat_int_all #mixed_sub

# List of your genes -> from paper and manually added
genes_of_interest <- c(
  "POU5F1", "NANOG", "THRB", "MKI67","MKI67", "TOP2A", "CDK1", "CCNB1", "PCNA", "RRM2", 
  "HAND2", "BAMBI", "LUM", "DCN", "DLK1",
  "ETV2", "MAML1", "PITX1", "SPRY1", "TRPS1", "SNAI2", "PDGFRB", "CSPG4", 
  "ANGPT2", "MATN4", "MIXL1", "DLL3", "SULF1", "COL4A1", "LOX", "KRT19", 
  "WNT5A", "NID2", "POSTN", "IGFBP3", "COL5A1", "HES2", "HRH1", "SMAD3", 
  "RGCC", "PLAUR", "IGFBP5", "NR3C1", "GATA6", "SGK1", "DKK1", "CRABP1", 
  "HEY1", "DLL4", "VWF", "CLDN5", "CXCR4", "GJA4", "GJA5", "SEMA3G", 
  "MCAM", "WFDC2", "DLL1", "ARL4A", "ZFPM2", "HES1", "PRDM1", "FOXO1", 
  "ESM1", "MECOM", "PECAM1", "ADM", "SOX18", "ERG", "PDGFB", "MMRN1", 
  "CDH5", "DUSP23", "CFD", "SOX7", "KDR", "ROBO4", "PLVAP", "FLI1", 
  "HHEX", "GATA2", "ELK3", "LEF1", "CPE", "EPCAM", "HES4", "PDGFA", 
  "RGS5", "PGF", "HES5", "MYC", "RBPJ", "EP300", "SPINT2", "ZEB2", 
  "EBF1", "NR2F2", "NRP2", "APLNR", "ZEB1", "ADAMTS18", "NTS", "NEFH", 
  "LY6H", "RGS16", "HAND1", "KCNE3", "TFPI2", "MESP1", "MESP2", "PTPRC", 
  "ALDH1A1", "CTGF", "FOXQ1", "PAX3", "AQP4", "ANKRD1", "TAGLN", "ACTA2"
)

#Keep only genes that exist in your Seurat object and get average expression per cluster 
genes_present <- intersect(genes_of_interest, rownames(seurat_obj))
avg_expr <- AverageExpression(seurat_obj, features = genes_present, group.by = "seurat_clusters", slot = "data")
avg_matrix <- as.matrix(avg_expr$RNA)
z_score <- t(scale(t(avg_matrix)))
colnames(z_score) <- gsub("g", "", colnames(z_score)) #"g0" becomes "0", for style
# Manually ordered to obtain a red diagonal
my_cluster_order <- c("7", "10", "3", "9", "4", "6", "5", "1", "2", "8", "0", "11", "12", "13", "14", "15", "16", "17", "18","19", "20") 
my_order <- intersect(my_cluster_order, colnames(z_score))
z_score_ordered <- z_score[, my_order]
var_sub <- apply(z_score_ordered, 1, var, na.rm = TRUE)
z_score_ordered <- z_score_ordered[var_sub > 0 & !is.na(var_sub), ]

# heatmap
h <- pheatmap(z_score_ordered,
         scale = "none",
         cluster_rows = TRUE,       # Keeps your custom order
         cluster_cols = FALSE, 
         treeheight_row = 0,        # Deletes the row dendrogram
         treeheight_col = 0,        # Deletes the column dendrogram (if any)
         fontsize_row = 5,          # Prevents name overlap
         height = 2,
         angle_col = 0,             # Keeps cluster names horizontal
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "Z-scored gene expression per cluster ")

# Save pdf
ggsave("heatmap.pdf", plot = h, width = 12, height = 8, dpi=600,  bg = "white")

# Printing cluster's marker gene to ask LLMs which tissue corresponds
genes_main <- apply(z_score_ordered, 2, function(x) {
names(x[x > 2]) })  #Filter Z-score > 2
genes_minor <- apply(z_score_ordered, 2, function(x) {
names(x[x > 1 & x < 2]) }) #Filter z-score > 1 & z-score < 2

# Print main and minor genes per cluster -> ASK LLMs to get cell type and cell state (if inferable)
print(genes_main)
print(genes_minor)

```

Cell Annotation Assignment on Whole integration
```{r, include=TRUE, eval=TRUE}
seurat_int_all <- readRDS("seurat_integrated_r06.rds")

# --------------- cluster identity --------------- #
Idents(seurat_int_all) <- "seurat_clusters"
new_cluster_ids <- c(
  "0" = "Fibroblast-like Mural",
  "1" = "Epicardial Progenitors",
  "2" = "Mural Progenitors", #Undifferentiated perivascular cells.
  "3" = "Epicardial-like",
  "4" = "Contractile Mural", #Pericytes/SMCs
  "5" = "Neural Crest / Pluripotent", #mixed - CRABP1, PAX3 (Neural Crest) + POU5F1, MIXL1 (Pluripotent/Early).
  "6" = "Neural Crest",
  "7" = "Venous / Capillary Endothelial",
  "8" = "Proliferating Neural Crest",
  "9" = "Proliferating Early Mesoderm",
  "10" = "Pan-Endothelial / Epithelial", #Highly Mixed. Contains pure Arterial ECs and Epithelial cells.
  "11" = "Neuromesoderm/Myeloid"
)
seurat_int_all <- RenameIdents(seurat_int_all, new_cluster_ids)
seurat_int_all$cell_type <- Idents(seurat_int_all)

# plots with annotation
DimPlot(seurat_int_all, reduction = "umap", label = FALSE) + ggtitle("Day 21 Clusters")
```

Some problems/mixed clusters:
Cluster 10 is a major issue: It mixes your most important Arterial ECs with EPCAM+ Epithelial cells. At higher resolution, these must separate.
Cluster 11 is a "collision": Neuromesoderm (MESP2) and Myeloid (PTPRC) cells should never be in the same cluster.
Cluster 5 is unresolved: It currently groups early pluripotent cells (POU5F1) with neural crest cells (CRABP1).
Next Step: Increase resolution to 0.8. If Cluster 10 and 11 still don't split, extract just those cells and perform sub-clustering on them specifically.


# Sub-clustering 10,11
```{r}
# 1. Reset and Subset
# Use the original numeric idents if you haven't renamed, 
# OR use the names if you have. This version is safe:
Idents(seurat_int_all) <- "seurat_clusters" 
mixed_sub <- subset(seurat_int_all, idents = c("10", "11"))

# 2. Switch to RNA Assay to find ALL your genes
# This fixes the "39 features not found" warning
DefaultAssay(mixed_sub) <- "RNA"

# 3. Define and Clean Features
genes_of_interest <- c("POU5F1", "NANOG", "THRB", "HAND2", "BAMBI", "LUM", "DCN", "DLK1", "ETV2", "MAML1", "PITX1", "SPRY1", "TRPS1", "SNAI2", "PDGFRB", "CSPG4", "ANGPT2", "MATN4", "MIXL1", "DLL3", "SULF1", "COL4A1", "LOX", "KRT19", "WNT5A", "NID2", "POSTN", "IGFBP3", "COL5A1", "HES2", "HRH1", "SMAD3", "RGCC", "PLAUR", "IGFBP5", "NR3C1", "GATA6", "SGK1", "DKK1", "CRABP1", "HEY1", "DLL4", "VWF", "CLDN5", "CXCR4", "GJA4", "GJA5", "SEMA3G", "MCAM", "WFDC2", "DLL1", "ARL4A", "ZFPM2", "HES1", "PRDM1", "FOXO1", "ESM1", "MECOM", "PECAM1", "ADM", "SOX18", "ERG", "PDGFB", "MMRN1", "CDH5", "DUSP23", "CFD", "SOX7", "KDR", "ROBO4", "PLVAP", "FLI1", "HHEX", "GATA2", "ELK3", "LEF1", "CPE", "EPCAM", "HES4", "PDGFA", "RGS5", "PGF", "HES5", "MYC", "RBPJ", "EP300", "SPINT2", "ZEB2", "EBF1", "NR2F2", "NRP2", "APLNR", "ZEB1", "ADAMTS18", "NTS", "NEFH", "LY6H", "RGS16", "HAND1", "KCNE3", "TFPI2", "MESP1", "MESP2", "PTPRC", "ALDH1A1", "CTGF", "FOXQ1", "PAX3", "AQP4", "ANKRD1", "TAGLN", "ACTA2")
cycle_genes <- c("MKI67", "TOP2A", "CDK1", "CCNB1", "PCNA", "RRM2")
targeted_features <- setdiff(genes_of_interest, cycle_genes)
targeted_features <- intersect(targeted_features, rownames(mixed_sub))

# 4. Re-Process with Scaling (The crucial fix)
mixed_sub <- NormalizeData(mixed_sub)
mixed_sub <- ScaleData(mixed_sub, features = targeted_features)
mixed_sub <- RunPCA(mixed_sub, features = targeted_features, verbose = FALSE)

# 5. Find Clusters (Resolution 1.3 is the "Sweet Spot" for 6 clusters)
mixed_sub <- FindNeighbors(mixed_sub, dims = 1:20)
mixed_sub <- FindClusters(mixed_sub, resolution = 1.2)
mixed_sub <- RunUMAP(mixed_sub, dims = 1:20)

# 6. Apply the 6-way naming
# Note: If it finds a 7th cluster (6), I added a placeholder name
new_sub_names <- c(
  "0" = "Proliferating Early Mesoderm", 
  "1" = "Arterial EC", #no EPCAM pure arterial
  "2" = "Contractile Mural",  
  "3" = "Mural Progenitors",  
  "4" = "Venous / Capillary Endothelial", 
  "5" = "Proliferating Myeloid"
)

mixed_sub <- RenameIdents(mixed_sub, new_sub_names)
mixed_sub$refined_type <- Idents(mixed_sub)

# 7. Visualize Result
DimPlot(mixed_sub, reduction = "umap", label = FALSE, repel = TRUE) + 
  ggtitle("Successfully Resolved: 10 & 11 Sub-types")
```

Sub-clustering 5
```{r}
# Subset Cluster 5
cluster5_sub <- subset(seurat_int_all, idents = "5")
DefaultAssay(cluster5_sub) <- "RNA"

# Re-scale and PCA using your existing targeted_features
cluster5_sub <- NormalizeData(cluster5_sub)
cluster5_sub <- ScaleData(cluster5_sub, features = targeted_features)
cluster5_sub <- RunPCA(cluster5_sub, features = targeted_features, verbose = FALSE)

# Find Sub-clusters (Try lower resolution first)
cluster5_sub <- FindNeighbors(cluster5_sub, dims = 1:20)
cluster5_sub <- FindClusters(cluster5_sub, resolution = 0.7)
cluster5_sub <- RunUMAP(cluster5_sub, dims = 1:20)


new_sub_5 <- c(
  "0" = "Pluripotent Progenitors", 
  "1" = "Venous / Capillary Endothelial", 
  "2" = "Endothelial Progenitors",  
  "3" = "Arterial EC", # This should be "Arterial EC (Proliferating)" but I prefer to not separate them
  "4" = "Arterial EC"
)

cluster5_sub <- RenameIdents(cluster5_sub, new_sub_5)
cluster5_sub$refined_type <- Idents(cluster5_sub)

# Visualize Result
DimPlot(cluster5_sub, reduction = "umap", label = FALSE, repel = TRUE) + 
  ggtitle("Successfully Resolved: 5 Sub-types")
```

Merge all sub-cluster to get one final_type Annotation 
```{r}
#---------- MERGE SUB-CLUSTERS with MAIN CLUSTERS ---------- #
# Initialize final types with your current main annotations
seurat_int_all$final_type <- as.character(seurat_int_all$cell_type)

# Inject refined names from sub-clusters 10 & 11
seurat_int_all$final_type[Cells(mixed_sub)] <- as.character(Idents(mixed_sub))

# Inject refined names from sub-cluster 5
seurat_int_all$final_type[Cells(cluster5_sub)] <- as.character(Idents(cluster5_sub))

# Set the new names as the active identity
Idents(seurat_int_all) <- "final_type"
#------------------------------------------------------------#

#Tidying the object - style
# Write out all 13 names in the exact sequence you want them to appear
my_custom_order <- c(
  "Arterial EC",
  "Venous / Capillary Endothelial",
  "Contractile Mural",
  "Fibroblast-like Mural",
  "Epicardial-like",
  "Neural Crest",
  "Pluripotent Progenitors",
  "Endothelial Progenitors",
  "Mural Progenitors",
  "Epicardial Progenitors",
  "Proliferating Neural Crest",
  "Proliferating Early Mesoderm",
  "Proliferating Myeloid"
)

# Overwrite final_type as an ordered factor
seurat_int_all$final_type <- factor(seurat_int_all$final_type, levels = my_custom_order)

# Clean up old metadata columns
seurat_int_all$strata_cluster <- NULL
seurat_int_all$cell_type <- NULL

# Create numbered clusters (1 to 13) matching the custom order
seurat_int_all$seurat_clusters <- as.character(as.numeric(seurat_int_all$final_type))

# Convert the numbers into an ordered factor so they sort correctly (1, 2, 3...)
seurat_int_all$seurat_clusters <- factor(seurat_int_all$seurat_clusters, levels = as.character(1:13))

# Plot by Number
Idents(seurat_int_all) <- "seurat_clusters"
DimPlot(seurat_int_all, reduction = "umap", label = TRUE, repel = TRUE) + 
  ggtitle("Main UMAP: Perfectly Ordered Numbers")

# Plot by Name
Idents(seurat_int_all) <- "final_type"
DimPlot(seurat_int_all, reduction = "umap", label = FALSE) + 
  ggtitle("Main UMAP: Ordered by Name")

# Save Final Object
#saveRDS(seurat_int_all, file="seurat_int_all.rds")
```

