---
title: "part1"
output: html_document
date: "2026-05-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(Seurat)
library(Matrix)
library(patchwork)
library(dplyr)
library(ggplot2)

options(Seurat.object.assay.version = "v3")
```

```{r}
# Read the files
counts <- readMM("counts.mtx")
meta <- read.table("meta.tsv")
features <- read.table("features.tsv")[,1]
dimnames(counts) <- list(features, rownames(meta))
seurat <- CreateSeuratObject(counts = counts, meta.data = meta)

# How many cells and genes?
cat("Genes:", nrow(counts), "\n")
cat("Cells:", ncol(counts), "\n")

# Look at your metadata 
table(meta$sample, meta$condition)

``` 
Split by sample -> 4 sample/seurat
```{r}
seurat_list <- SplitObject(seurat, split.by = "sample")
```

Quality Control (QC) and filtering low-quality cells
```{r}
#Plotting "nFeature_RNA", "nCount_RNA", "percent.mt"
seurat_list <- lapply(seurat_list, function(x) {
  x[["percent.mt"]] <- PercentageFeatureSet(x, pattern = "^MT[-\\.]")
  x
})
lapply(seurat_list, function(x) {
  VlnPlot(x, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
})
lapply(seurat_list, function(x) {
  plot1 <- FeatureScatter(x, feature1 = "nCount_RNA", feature2 = "percent.mt") + ggtitle(unique(x$sample))
  plot2 <- FeatureScatter(x, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + ggtitle(unique(x$sample))
  plot1 + plot2
})

#Filter the cells by fixed threshold and find most 3000 differentially expressed genes
seurat_list <- lapply(seurat_list, function(x) {
  x <- subset(x, subset = nFeature_RNA > 500 & nFeature_RNA < 3500 & percent.mt < 10)
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, nfeatures = 3000)
  x <- ScaleData(x, vars.to.regress = c("nFeature_RNA", "percent.mt"))
  x
})
```

PCA over 4 samples
```{r}
seurat_list <- lapply(seurat_list, function(x) {
  x <- RunPCA(x, npcs = 50)
  x
})

#2 method to visualize data
#Elbow PCA
lapply(seurat_list, function(x) {
  ElbowPlot(x, ndims = ncol(Embeddings(x, "pca")))
})

#Heatmap PCA
lapply(seurat_list, function(x) {
  PCHeatmap(x, dims = 1:15, cells = 500, balanced = TRUE, ncol = 4)
})
```

Data Integration: YES or NO 
Is there a batch effects ? If cells of the same known type (e.g., progenitors or neurons/genes signature) from different replicates (only between WT or betw KO) fail to intermix on a UMAP or t-SNE plot, a batch effect is likely present.
```{r}
# Merge all samples without integration
seurat_merged <- merge(x = seurat_list$B1_17_1_day21, 
                       y = c(seurat_list$D2_22_3_day21, 
                             seurat_list$wt_2_day21, 
                             seurat_list$wt_4_day21),
                       add.cell.ids = c("B1", "D2", "wt2", "wt4")) %>%
  NormalizeData() %>%
  FindVariableFeatures(nfeatures = 3000) %>%
  ScaleData(vars.to.regress = c("nFeature_RNA", "percent.mt")) %>%
  RunPCA(npcs = 50) %>%
  RunUMAP(dims = 1:15)

# Plot by sample
DimPlot(seurat_merged, group.by = "sample")

# Plot by condition
DimPlot(seurat_merged, group.by = "condition")

# Subset for WT and plot
DimPlot(subset(seurat_merged, subset = condition == "WT"), 
                group.by = "sample", 
                pt.size = 0.5) + ggtitle("Wild Type Samples")

# Subset for KO and plot
DimPlot(subset(seurat_merged, subset = condition == "KO"), 
                group.by = "sample", 
                pt.size = 0.5) + ggtitle("KO Samples")

# Subset the object to only include Wild Type samples or KO
seurat_wt <- subset(seurat_merged, subset = condition == "WT")
seurat_ko <- subset(seurat_merged, subset = condition == "KO")

# --- WILD TYPE (WT) SAMPLES ---
FeaturePlot(seurat_wt, features = "POU5F1", split.by = "sample")
FeaturePlot(seurat_wt, features = "MIXL1", split.by = "sample")
FeaturePlot(seurat_wt, features = "EOMES", split.by = "sample")
FeaturePlot(seurat_wt, features = "HAND1", split.by = "sample")
FeaturePlot(seurat_wt, features = "MKI67", split.by = "sample")
FeaturePlot(seurat_wt, features = "ETV2", split.by = "sample")
FeaturePlot(seurat_wt, features = "TFPI2", split.by = "sample")
FeaturePlot(seurat_wt, features = "NRP2", split.by = "sample")
FeaturePlot(seurat_wt, features = "KDR", split.by = "sample")
FeaturePlot(seurat_wt, features = "CLDN5", split.by = "sample")
FeaturePlot(seurat_wt, features = "PECAM1", split.by = "sample")
FeaturePlot(seurat_wt, features = "GJA4", split.by = "sample")
FeaturePlot(seurat_wt, features = "GJA5", split.by = "sample")
FeaturePlot(seurat_wt, features = "CXCR4", split.by = "sample")
FeaturePlot(seurat_wt, features = "NR2F2", split.by = "sample")
FeaturePlot(seurat_wt, features = "APLNR", split.by = "sample")
FeaturePlot(seurat_wt, features = "PRCP", split.by = "sample")
FeaturePlot(seurat_wt, features = "DCN", split.by = "sample")
FeaturePlot(seurat_wt, features = "DLK1", split.by = "sample")
FeaturePlot(seurat_wt, features = "SULF1", split.by = "sample")
FeaturePlot(seurat_wt, features = "COL4A1", split.by = "sample")
FeaturePlot(seurat_wt, features = "ACTA2", split.by = "sample")
FeaturePlot(seurat_wt, features = "LUM", split.by = "sample")
FeaturePlot(seurat_wt, features = "PDGFRB", split.by = "sample")
FeaturePlot(seurat_wt, features = "CSPG4", split.by = "sample")

# --- KNOCKOUT (KO) SAMPLES ---
FeaturePlot(seurat_ko, features = "POU5F1", split.by = "sample")
FeaturePlot(seurat_ko, features = "MIXL1", split.by = "sample")
FeaturePlot(seurat_ko, features = "EOMES", split.by = "sample")
FeaturePlot(seurat_ko, features = "HAND1", split.by = "sample")
FeaturePlot(seurat_ko, features = "MKI67", split.by = "sample")
FeaturePlot(seurat_ko, features = "ETV2", split.by = "sample")
FeaturePlot(seurat_ko, features = "TFPI2", split.by = "sample")
FeaturePlot(seurat_ko, features = "NRP2", split.by = "sample")
FeaturePlot(seurat_ko, features = "KDR", split.by = "sample")
FeaturePlot(seurat_ko, features = "CLDN5", split.by = "sample")
FeaturePlot(seurat_ko, features = "PECAM1", split.by = "sample")
FeaturePlot(seurat_ko, features = "GJA4", split.by = "sample")
FeaturePlot(seurat_ko, features = "GJA5", split.by = "sample")
FeaturePlot(seurat_ko, features = "CXCR4", split.by = "sample")
FeaturePlot(seurat_ko, features = "NR2F2", split.by = "sample")
FeaturePlot(seurat_ko, features = "APLNR", split.by = "sample")
FeaturePlot(seurat_ko, features = "PRCP", split.by = "sample")
FeaturePlot(seurat_ko, features = "DCN", split.by = "sample")
FeaturePlot(seurat_ko, features = "DLK1", split.by = "sample")
FeaturePlot(seurat_ko, features = "SULF1", split.by = "sample")
FeaturePlot(seurat_ko, features = "COL4A1", split.by = "sample")
FeaturePlot(seurat_ko, features = "ACTA2", split.by = "sample")
FeaturePlot(seurat_ko, features = "LUM", split.by = "sample")
FeaturePlot(seurat_ko, features = "PDGFRB", split.by = "sample")
FeaturePlot(seurat_ko, features = "CSPG4", split.by = "sample")
```

Data Integration
```{r}
counts_int <- readMM("counts.mtx")
meta_int <- read.table("meta.tsv")
features_int <- read.table("features.tsv")[,1]
dimnames(counts_int) <- list(features_int, rownames(meta_int))
seurat_int <- CreateSeuratObject(counts = counts_int, meta.data = meta_int)
#Splitting dataset
seurat_list_int <- SplitObject(seurat_int, split.by = "sample") #KO vs WT
```

Filtering
```{r}
seurat_list_int <- lapply(seurat_list_int, function(x) {
  x[["percent.mt"]] <- PercentageFeatureSet(x, pattern = "^MT[-\\.]")
  x
})

seurat_list_int <- lapply(seurat_list_int, function(x) {
  x <- subset(x, subset = nFeature_RNA > 500 & nFeature_RNA < 3500 & percent.mt < 10)
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, nfeatures = 3000)
  x
})
```

FindIntegrationAnchors
```{r}
seurat_list_KO = seurat_list_int[c("B1_17_1_day21", "D2_22_3_day21")]
seurat_list_WT = seurat_list_int[c("wt_2_day21", "wt_4_day21")]
anchorsKO <- FindIntegrationAnchors(object.list = seurat_list_KO, dims = 1:15)
anchorsWT <- FindIntegrationAnchors(object.list = seurat_list_WT, dims = 1:15)
```

IntegrateData
```{r}
seurat_int_KO <- IntegrateData(anchorsKO, dims = 1:15)
seurat_int_WT <- IntegrateData(anchorsWT, dims = 1:15)
```

```{r}
seurat_int_KO <- ScaleData(seurat_int_KO)
seurat_int_KO <- RunPCA(seurat_int_KO, npcs = 15)
seurat_int_KO <- RunUMAP(seurat_int_KO, dims = 1:15)
seurat_int_KO <- RunTSNE(seurat_int_KO, dims = 1:15)
seurat_int_KO <- FindNeighbors(seurat_int_KO, dims = 1:15) %>% FindClusters(resolution = 0.6)
```

```{r}
seurat_int_WT <- ScaleData(seurat_int_WT)
seurat_int_WT <- RunPCA(seurat_int_WT, npcs = 15)
seurat_int_WT <- RunUMAP(seurat_int_WT, dims = 1:15)
seurat_int_WT <- RunTSNE(seurat_int_WT, dims = 1:15)
seurat_int_WT <- FindNeighbors(seurat_int_WT, dims = 1:15) %>% FindClusters(resolution = 0.6)
```

```{r}
DefaultAssay(seurat_int_WT) <- "integrated" #RNA for analysis (real counts) and integrated for visualization (corrected counts)
plot1 <- UMAPPlot(seurat_int_WT, group.by="sample")
plot2 <- UMAPPlot(seurat_int_WT, label = T)

# Use the real RNA expression values
DefaultAssay(seurat_int_WT) <- "RNA"
# FeaturePlot will now pull expression from the RNA assay,
# but the UMAP layout is the one computed from the integrated assay
plot3 <- FeaturePlot(seurat_int_WT, c("PECAM1","CD34","CLDN5"), ncol = 2, pt.size = 0.1)
((plot1 / plot2) | plot3) + plot_layout(width = c(1,2))
```

```{r}
DefaultAssay(seurat_int_KO) <- "integrated"
plot1 <- UMAPPlot(seurat_int_KO, group.by="sample")
plot2 <- UMAPPlot(seurat_int_KO, label = T)

# Use the real RNA expression values
DefaultAssay(seurat_int_WT) <- "RNA"
# FeaturePlot will now pull expression from the RNA assay,
# but the UMAP layout is the one computed from the integrated assay
plot3 <- FeaturePlot(seurat_int_KO, c("PECAM1","CD34","CLDN5"), ncol = 2, pt.size = 0.1)
((plot1 / plot2) | plot3) + plot_layout(width = c(1,2))
```

```{r}
saveRDS(seurat_int_KO, file="seurat_int_KO.rds")
saveRDS(seurat_int_WT, file="seurat_int_WT.rds")
```

Cluster Annotation
```{r}
DefaultAssay(seurat_int_WT) <- "RNA"
DefaultAssay(seurat_int_KO) <- "RNA"

ct_markers <- c(
  "POU5F1", # pluripotent stem cells
  "MIXL1", "EOMES", "HAND1", # mesoderm progenitors
  "MKI67", # cell cycle / proliferation
  "ETV2", # early endothelial progenitors
  "TFPI2", "NRP2", # late endothelial progenitors
  "KDR", "CLDN5", "PECAM1", # mature endothelial cells
  "GJA4", "GJA5", "CXCR4", # arterial-like endothelial
  "NR2F2", "APLNR", "PRCP", # venous-like endothelial
  "DCN", "DLK1", "SULF1", "COL4A1", # general mural cells
  "ACTA2", "LUM", # vascular smooth muscle cells
  "PDGFRB", "CSPG4" # pericytes
)

DoHeatmap(seurat_int_WT, features = ct_markers) + guides(colour=FALSE)
DoHeatmap(seurat_int_KO, features = ct_markers) + guides(colour=FALSE)
```


